Parallel Simulation of Adaptive Random Boolean Networks

### 2016-07-14 Paper/paper.tex
Comment out last paragraph in 3.II as it repeats 3.III + cosmetics

\paragraph{II. Updating states of nodes in parallel.} Another way to reduce an execution time of modeling is to parallelize calculations of nodes' states at a particular iteration. In this study, we propose a GPGPU algorithm which speeds up sequential heuristics of type (I). As it does not influence the logic of a search, this algorithm can be used in a combination with any sequential heuristic. The main idea of the algorithm is to update states of different nodes in a network in parallel on distinct Graphic Processor Units (GPUs). More formally,...

	[To do. A description of parallel algorithm]

@@

	However, the lenghts of attractors and transient periods grow faster than a power law[][] when the size of the network increases. This leads to a proportional growth of state updates one has to perform to find an attractor. The number of state updates turned out to be more important to the algorithm performance due to the lags between GPU kernel launches and the need to transfer data back from the GPU.

@@

	Intuitively, Liu-Bassler algorithm of attractor search would have shown better performance than the Knuth's one, as the latter requires at least $3 \mu$ state updates to find an attractor of length $\mu$. But in order to use Liu-Bassler algorithm one needs an a-priori knowledge of attractor and transient period distributions in order to choose the checkpoints correctly. This distribution depends on a network size, connectivity and structure (e. g. modular networks tend to have longer (shorter?) attractors than non-modular ones [ref]). We obtained an empirical distribution of attractor and transient lengths for different kinds of networks in our preliminary experiments with Knuth's algorithm. The checkpoints $T=[100, 200, 1000, 2000, ... \mu_{max}, 2 \mu_{max}]$ seem to be the best trade-off between the number of state updates and the percentage of attractors found. However, with these checkpoints the difference in the number of state updates between the two algorithms is negligible. We used Knuth's algorithm in our future experiments, as it is more general and requires no a-priori knowledge of the network's behaviour.

	Both algorithms were implemented using C++ programming language. The implementation is cross-platform and requires a GPU with CUDA compute capability at least 1.2. The experiments were conducted on a cluster consisting of 20 machines with NVidia GeForce GT 640 graphic cards. The results presented below are averaged over 10 realizations of the evolution process.

	\begin{figure}[h!]
		\centering
		\includegraphics[width=0.7\textwidth]{plots/evolution}
		\caption{Evolution of mean network connectivity}
	\end{figure}

	\begin{figure}[h!]
		\centering
		\includegraphics[width=0.7\textwidth]{plots/n_rew}
		\caption{Evolution of mean connectivity for different number of nodes rewired each epoch; $N=1000$}
	\end{figure}

	\begin{figure}[h!]
		\centering
		\includegraphics[width=0.7\textwidth]{plots/attractors}
		\caption{Empirical distribution of attractor lengths}
	\end{figure}

>>>>

\paragraph{II. Updating states of nodes in parallel.} Another way to reduce an execution time of modeling is to parallelize calculations of nodes' states at a particular iteration. In this study, we propose a GPGPU algorithm which speeds up sequential heuristics of type (I). As it does not influence the logic of a search, this algorithm can be used in a combination with any sequential heuristic. The main idea of the algorithm is to update states of different nodes in a network in parallel on distinct Graphic Processor Units (GPUs). More formal description is presented in Algorithm \ref{algo:gpu}.

	\begin{algorithm}[ht!]
		\label{algo:gpu}
		GenerateNetwork()\;
		$epoch \leftarrow 0$\;
		\While{ $epoch < epochsCount$ }
		{
			$gpuNetwork \leftarrow$ ConvertNetworkToGpuRepresentation()\;
			ZeroInitialize($stateSum$)\;
			CopyToGpu($gpuNetwork$, $currentState$, $stateSum$)\;
			\While{attractor not found}
			{ 	
				UpdateStateOnGpu($currentState$, $stateSum$)\;
			}
			CopyFromGpu($stateSum$)\;
			$nodesToRewire \leftarrow$ ChooseRandomNodes($N_{rew}$)\;
			ActivityDependentRewiring($nodesToRewire$, $stateSum$)\;
			$epoch \leftarrow epoch + 1$\;
		}
		\caption{General algorithm of network evolution on GPU}
	\end{algorithm}

@@

	%However, the lenghts of attractors and transient periods grow %faster than a power law[][] when the size of the network %increases. This leads to a proportional growth of state updates %one has to perform to find an attractor. The number of state %updates turned out to be more important to the algorithm %performance due to the lags between GPU kernel launches and the %need to transfer data back from the GPU.

@@

	Intuitively, Liu-Bassler algorithm of attractor search would have shown better performance than the Knuth's one, as the latter requires at least $3 \mu$ state updates to find an attractor of length $\mu$. But in order to use Liu-Bassler algorithm one needs to choose the checkpoints correctly, which requires an a-priori knowledge of attractor and transient period distributions. This distribution depends on a network size, connectivity and structure (e. g. modular networks tend to have longer (shorter?) attractors than non-modular ones [ref]). We obtained an empirical distribution of attractor and transient lengths for different kinds of networks in our preliminary experiments with Knuth's algorithm. The checkpoints $T=[100, 200, 1000, 2000, ... \mu_{max}, 2 \mu_{max}]$ seem to be the best trade-off between the number of state updates and the percentage of attractors found. However, with these checkpoints the difference in the number of state updates between the two algorithms is negligible. We used Knuth's algorithm in our future experiments, as it is more general and requires no a-priori knowledge of the network's behaviour.

	Both algorithms were implemented using C++ programming language. The implementation is cross-platform and requires a GPU with CUDA compute capability at least 1.2.

	To test the applicability of the approaches presented in Section \ref{sect:algorithm}, a set of experiments was performed. The experiments were conducted on a cluster consisting of 20 nodes with NVidia GeForce GT 640 graphic cards. The results below were obtained by averaging over 10 realizations of the evolution process.

	\begin{figure}[ht!]
		\begin{minipage}[t]{0.45\textwidth}
			\includegraphics[width=1.0\textwidth]{plots/evolution}
			\caption{Evolution of mean network connectivity}
		\end{minipage}\hfill
		\begin{minipage}[t]{0.45\textwidth}
			\includegraphics[width=1.0\textwidth]{plots/n_rew}
			\caption{Evolution of mean connectivity for different number of nodes rewired each epoch; $N=1000$}
		\end{minipage}
	\end{figure}

	\begin{figure}[ht!]
		\begin{minipage}[t]{0.45\textwidth}
			\centering
			\includegraphics[width=1.0\textwidth]{plots/attractors}
			\caption{Empirical distribution of attractor lengths}
		\end{minipage}\hfill
		\begin{minipage}[t]{0.45\textwidth}
			\includegraphics[width=1.0\textwidth]{plots/speedup}
			\caption{Execution time of the GPU algorithm compared to the serial one}
		\end{minipage}
	\end{figure}

	\begin{table}[ht!]
		\begin{tabular}{r | c c c c c}
			& \multicolumn{5}{c}{Nodes count} \\
			$\mu_{max}$	& 100	& 	200	&	300	&	500	&	750\\ \hline
			$5 \cdot 10^4$	&	99.77	&	88.19	&	71.19	&	45.48	&	29.62\\
			$2.5 \cdot 10^5$	&	99.99	&	95.92	&	81.24	&	56.10	&	35.96\\
			$5 \cdot 10^5$	&	100.00	&	97.43	&	86.58	&	59.36	&	38.73\\
			$2.5 \cdot 10^6$	&	100.00	&	99.14	&	92.99	&	68.74	&	46.46\\
			$5 \cdot 10^6$	&	100.00	&	99.53	&	94.40	&	73.00	&	49.22\\
			$1 \cdot 10^7$	&	100.00	&	99.75	&	96.08	&	77.73	&	53.81\\
		\end{tabular}
		\caption{Percentage $R$ of attractors found during the evolution process with attractor length limeted by $\mu_{max}$}
	\end{table}
	\begin{table}[ht!]
		\begin{tabular}{r | c c c c c}
			& \multicolumn{5}{c}{Nodes count} \\
			$\mu_{max}$	& 100	& 	200	&	300	&	500	& 750\\
			\hline
			$5 \cdot 10^4$	&	2.64	&	2.59	&	2.58	&	2.52	&	2.47\\
			$2.5 \cdot 10^5$	&	2.66	&	2.60	&	2.56	&	2.51	&	2.48\\
			$5 \cdot 10^5$	&	2.68	&	2.59	&	2.55	&	2.51	&	2.48\\
			$2.5 \cdot 10^6$	&	2.69	&	2.60	&	2.55	&	2.51	&	2.48\\
			$5 \cdot 10^6$	&	2.69	&	2.59	&	2.55	&	2.52	&	2.48\\
			$1 \cdot 10^7$	&	2.68	&	2.62	&	2.55	&	2.51	&	2.48\\
		\end{tabular}
		\caption{Mean steady state connectivity $K_{ss}$ obtained as a result of evolution process with attractor length limeted by $\mu_{max}$}
	\end{table}
